---
title: "Kirstie_Regression"
author: "Kirstie Turnbull"
date: "11/29/2020"
output: html_document
---

Goal is to run this model: "log(E(cases_t)) = offset(log(population))+ mobility_t + mask_use_t + demographics + political party + spline(t)" OR "Cases ~ bla + bs(time, knots = one knot every 30 days)" instead of "spline(t)"

```{r}
library(tidyverse)
library(lubridate)
library(rvest)
library(stringr)
library(zoo)
library(directlabels)
```


```{r}
# population data (same as Anja's)
url <- "https://en.wikipedia.org/wiki/List_of_states_and_territories_of_the_United_States_by_population"
h <- read_html(url)
```

```{r}
# looking for the first table on the page
tab <- h %>% html_nodes("table") %>% .[1] %>% html_table(fill = TRUE) %>% .[[1]]
```

```{r}
# narrow to state/population
pop_states <- tab %>% select("State","Census population")
# delete extra rows from wikipedia
pop_states <- pop_states[2:57,]
```

```{r}
# fix variable types of population data
pop_states$State <- as.factor(pop_states$State)
pop_states$`Census population` <- as.numeric(str_replace_all(pop_states$`Census population`, ",", ""))
# do we even want Guam, American Samoa, etc.?
```

```{r}
# fix variable name for population data
colnames(pop_states) <- c("state","population")
```

Okay I think this is good enough for population data for now!

Now for mobility data
```{r}
# read it in!
mobility <- read_csv("2020_US_Region_Mobility_report.csv")
```

```{r}
mobility <- mobility %>% filter(is.na(sub_region_2)==TRUE)
```

```{r}
mobility <- mobility %>% filter(is.na(sub_region_1)==FALSE)
```

```{r}
# update state column name
colnames(mobility)[3] <- "state"
```

```{r}
# update state variable type
mobility$state <- as.factor(mobility$state)
```

Okay, that's enough for mobility data, now for COVID data:
```{r}
# same as Anja's!
url <- "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv"
covid <- read_csv(url)
```

```{r}
# make state a factor
covid$state <- as.factor(covid$state)
```

Still need governor's political party/demographic/mask-use data before regression!

```{r}
# political party
url <- "https://en.wikipedia.org/wiki/List_of_current_United_States_governors"
h <- read_html(url)
```

```{r}
# looking for the first table on the page
tab <- h %>% html_nodes("table") %>% .[1] %>% html_table(fill = TRUE) %>% .[[1]]
```

```{r}
# remove unnecessary rows
gov_political <- tab[2:51,1:5]
```

```{r}
# fix column names
colnames(gov_political)[1] <- "state"
colnames(gov_political)[5] <- "gov_party"
```

```{r}
# narrow to only the columns we care about
gov_political <- gov_political %>% select("state","gov_party")
```

```{r}
# fix weird entries for Minnesota and West Virginia
gov_political[23,2] <- "Democratic"
gov_political[48,2] <- "Republican"
```

```{r}
# make state and political party factors 
gov_political$state <- as.factor(gov_political$state)
gov_political$gov_party <- as.factor(gov_political$gov_party)
```

Done with governor's political party! Now taking a look at mask data/demographic data

Demogs 
```{r}
url <- "https://en.wikipedia.org/wiki/Demographics_of_the_United_States"
h <- read_html(url)
```

```{r}
# looking for the first table on the page
tab <- h %>% html_nodes("table") %>% .[28] %>% html_table(fill = TRUE) %>% .[[1]]
```

```{r}
racial_demog <- tab
```

```{r}
racial_demog$`State or territory` <- as.factor(racial_demog$`State or territory`)
racial_demog$`Black orAfrican American` <- str_replace_all(racial_demog$`Black orAfrican American`, "%", "")
racial_demog$`Black orAfrican American` <- as.numeric(racial_demog$`Black orAfrican American`)
racial_demog$`American Indianand Alaska Native` <- str_replace_all(racial_demog$`American Indianand Alaska Native`, "%", "")
racial_demog$`American Indianand Alaska Native` <- as.numeric(racial_demog$`American Indianand Alaska Native`)
```

```{r}
mask <- read.csv("covidcast-fb-survey-smoothed_wearing_mask-2020-04-06-to-2020-12-01.csv")
```

# recode states
```{r}
list <- c("ak"="Alaska","al"="Alabama","az"="Arizona","ar"="Arkansas","ca"="California","co"="Colorado","ct"="Connecticut","de"="Delaware","fl"="Florida","ga"="Georgia","hi"="Hawaii","ia"="Iowa","id"="Idaho","il"="Illinois","in"="Indiana","ks"="Kansas","ky"="Kentucky","la"="Louisiana","ma"="Massachusetts","md"="Maryland","me"="Maine","mi"="Michigan","mn"="Minnesota","mo"="Missouri","ms"="Mississippi","mt"="Montana","nc"="North Carolina","ne"="Nebraska","nh"="New Hampshire","nj"="New Jersey","nm"="New Mexico","nv"="Nevada","ny"="New York","oh"="Ohio","ok"="Oklahoma","or"="Oregon","pa"="Pennsylvania","sc"="South Carolina","tn"="Tennessee","tx"="Texas","ut"="Utah","va"="Virginia","wa"="Washington","wi"="Wisconsin","wv"="West Virginia","ri"="Rhode Island","sd"="South Dakota","nd"="North Dakota","vt"="Vermont","wy"="Wyoming")
```

```{r}
mask <- mask %>% mutate(state=recode(geo_value,!!!list))
```

```{r}
mask$state <- as.factor(mask$state)
```


# make dates type date
```{r}
mask$time_value <- ymd(mask$time_value)
```

```{r}
colnames(mask)[4] <- "date"
```



```{r}
colnames(racial_demog)[1] <- "state"
```

```{r}
join1 <- left_join(gov_political,pop_states,by="state")
```

```{r}
join2 <- left_join(join1,racial_demog,by="state")
```

```{r}
join3 <- left_join(mobility,join2,by="state")
```

```{r}
final_df <- left_join(join3,covid,by=c("state","date"))
```

```{r}
final_df <- left_join(mask,final_df,by=c("state","date"))
```

```{r}
final_df$time <- as.numeric(final_df$date - ymd("2020-09-07"))
```

```{r}
url <- "https://en.wikipedia.org/wiki/List_of_United_States_cities_by_population"
h <- read_html(url)
```

```{r}
tab <- h %>% html_nodes("table") %>% .[5] %>% html_table(fill = TRUE) %>% .[[1]]
```

```{r}
large_cities <- tab
```

```{r}
large_cities <- large_cities[1:50,3]
```

```{r}
large_cities <- as.factor(large_cities)
```

```{r}
large_cities <- as.data.frame(table(large_cities))
```

```{r}
colnames(large_cities) <- c("state","num_large_cities")
```

```{r}
large_cities <- large_cities %>% filter(state != "District of Columbia")
```

```{r}
url <- "https://en.wikipedia.org/wiki/List_of_states_and_territories_of_the_United_States_by_population_density"
h <- read_html(url)
# looking for the first table on the page
tab <- h %>% html_nodes("table") %>% .[2] %>% html_table(fill = TRUE) %>% .[[1]]
```

```{r}
pop_density <- tab
```

```{r}
pop_density <- pop_density[2:57,c(1,4)]
```

```{r}
colnames(pop_density)[1] <- "state"
colnames(pop_density)[2] <- "density"
```

```{r}
pop_density$density <- as.numeric(pop_density$density)
```

```{r}
final_df <- left_join(final_df,large_cities,by="state")
```

```{r}
final_df <- left_join(final_df,pop_density,by="state")
```

```{r}
final_df$num_large_cities <- str_replace_na(final_df$num_large_cities,replacement = 0)
```

```{r}
final_df$num_large_cities <- as.numeric(final_df$num_large_cities)
```

```{r}
final_df <- final_df %>% filter(date < ymd("2020-11-25"))
```


```{r}
summary(final_df)
```

```{r}
final_df$state <- as.factor(final_df$state)
```

```{r}
final_df <- final_df %>% filter(state != "dc")
```

```{r}
summary(final_df)
```

```{r}
final_df <- final_df %>% group_by(state) %>% mutate(new_cases = cases - lag(cases))
```

```{r}
train_set <- final_df %>% filter(time < 70)
test_set <- final_df %>% filter(time >= 70)
```

```{r}
scatter.smooth(final_df$value,log(final_df$cases),col="pink")
```

Looks like as percentage of people wearing masks in public increases, log(cases) increases and then decreases. We may want to add a polynomial term.

```{r}
scatter.smooth(final_df$retail_and_recreation_percent_change_from_baseline,log(final_df$cases),col="pink")
```

Looks like as retail mobility increases, log(cases) increases and then decreases. We may want to add a quadratic term.

```{r}
scatter.smooth(final_df$grocery_and_pharmacy_percent_change_from_baseline,log(final_df$cases),col="pink")
```

Looks like as grocery store mobility increases, log(cases) increases and then decreases. 

```{r}
scatter.smooth(final_df$parks_percent_change_from_baseline,log(final_df$cases),col="pink")
```

Looks like as parks mobility increases, log(cases) decreases.

```{r}
scatter.smooth(final_df$transit_stations_percent_change_from_baseline,log(final_df$cases),col="pink")
```

Looks like as transit mobility increases, log(cases) increases and then decreases. We may want to try a quadratic term. 

```{r}
scatter.smooth(final_df$workplaces_percent_change_from_baseline,log(final_df$cases),col="pink")
```

As workplace mobility increases, log(cases) does not appear to change much.

```{r}
scatter.smooth(final_df$residential_percent_change_from_baseline,log(final_df$cases),col="pink")
```

Looks like as residential mobility increases, log(cases) increases, but barely. 

```{r}
plot(final_df$gov_party,log(final_df$cases))
```

It looks like log(cases) is distributed pretty equally among democratic and republican states.

```{r}
scatter.smooth(log(final_df$population),log(final_df$cases),col="pink")
```

As log(population) increases, log(cases) increases. 

```{r}
scatter.smooth(log(final_df$`Black orAfrican American`),log(final_df$cases),col="pink")
```

As log(percentage of the population that is Black or African American) increases, log(cases) increases. 

```{r}
scatter.smooth(log(final_df$`American Indianand Alaska Native`),log(final_df$cases),col="pink")
```

As log(percentage of the population that is American Indian or Alaskan Native) increases, log(cases) decreases. 

```{r}
plot(final_df$num_large_cities,log(final_df$cases),col="pink")
```

Looks like as the number of cities in the top 50 larges cities by population increases, log(cases) increases. 

```{r}
scatter.smooth(log(final_df$density),log(final_df$cases),col="pink")
```

Looks like as log(population density) increases, log(cases) increases. 


```{r}
# just throw everything in
library(splines)
library(splines2)
mod1 <- lm(log(cases) ~ retail_and_recreation_percent_change_from_baseline + 
             grocery_and_pharmacy_percent_change_from_baseline + 
             parks_percent_change_from_baseline + 
             transit_stations_percent_change_from_baseline +
             workplaces_percent_change_from_baseline +
             residential_percent_change_from_baseline +
             value +
             gov_party + 
             density +
             num_large_cities +
             `Black orAfrican American` +
             `American Indianand Alaska Native` +
             bSpline(time,df=5),
             offset = log(population), 
           data=train_set)
summary(mod1)
```

```{r}
mod1_preds <- predict.lm(mod1,test_set)
plot(exp(mod1_preds),test_set$cases)
abline(0,1)
```

```{r}
sqrt(mean(((exp(mod1_preds) - test_set$cases)^2), na.rm=TRUE))
```

```{r}
# remove governor party
mod2 <- lm(log(cases) ~ retail_and_recreation_percent_change_from_baseline + 
             grocery_and_pharmacy_percent_change_from_baseline + 
             parks_percent_change_from_baseline + 
             transit_stations_percent_change_from_baseline +
             workplaces_percent_change_from_baseline +
             residential_percent_change_from_baseline +
             value +
             density +
             num_large_cities +
             `Black orAfrican American` +
             `American Indianand Alaska Native` +
             bSpline(time,df=5),
             offset = log(population), 
           data=train_set)
summary(mod2)
```

```{r}
mod2_preds <- predict.lm(mod2,test_set)
plot(exp(mod2_preds),test_set$cases)
abline(0,1)
```

```{r}
sqrt(mean(((exp(mod2_preds) - test_set$cases)^2), na.rm=TRUE))
```

```{r}
# remove grocery mobility
mod3 <- lm(log(cases) ~ retail_and_recreation_percent_change_from_baseline + 
             parks_percent_change_from_baseline + 
             transit_stations_percent_change_from_baseline +
             workplaces_percent_change_from_baseline +
             residential_percent_change_from_baseline +
             value +
             density +
             num_large_cities +
             `Black orAfrican American` +
             `American Indianand Alaska Native` +
             bSpline(time,df=5),
             offset = log(population), 
           data=train_set)
summary(mod3)
```

```{r}
mod3_preds <- predict.lm(mod3,test_set)
plot(exp(mod3_preds),test_set$cases)
abline(0,1)
```

```{r}
sqrt(mean(((exp(mod3_preds) - test_set$cases)^2), na.rm=TRUE))
# got worse! add it back
```

```{r}
# add quadratic term for retail
mod4 <- lm(log(cases) ~ retail_and_recreation_percent_change_from_baseline + 
             I(retail_and_recreation_percent_change_from_baseline^2) +
             grocery_and_pharmacy_percent_change_from_baseline + 
             parks_percent_change_from_baseline + 
             transit_stations_percent_change_from_baseline +
             workplaces_percent_change_from_baseline +
             residential_percent_change_from_baseline +
             value +
             density +
             num_large_cities +
             `Black orAfrican American` +
             `American Indianand Alaska Native` +
             bSpline(time,df=5),
             offset = log(population), 
           data=train_set)
mod4_preds <- predict.lm(mod4,test_set)
plot(exp(mod4_preds),test_set$cases)
abline(0,1)
sqrt(mean(((exp(mod4_preds) - test_set$cases)^2), na.rm=TRUE))
# worse for prediction
```

```{r}
# add a quadratic term for grocery mobility
mod5 <- lm(log(cases) ~ retail_and_recreation_percent_change_from_baseline + 
             grocery_and_pharmacy_percent_change_from_baseline + 
             I(grocery_and_pharmacy_percent_change_from_baseline^2) +
             parks_percent_change_from_baseline + 
             transit_stations_percent_change_from_baseline +
             workplaces_percent_change_from_baseline +
             residential_percent_change_from_baseline +
             value +
             density +
             num_large_cities +
             `Black orAfrican American` +
             `American Indianand Alaska Native` +
             bSpline(time,df=5),
             offset = log(population), 
           data=train_set)
mod5_preds <- predict.lm(mod5,test_set)
plot(exp(mod5_preds),test_set$cases)
abline(0,1)
sqrt(mean(((exp(mod5_preds) - test_set$cases)^2), na.rm=TRUE))
# better for prediction!
```

```{r}
# add transit mobility quadratic term
mod6 <- lm(log(cases) ~ retail_and_recreation_percent_change_from_baseline + 
             grocery_and_pharmacy_percent_change_from_baseline + 
             I(grocery_and_pharmacy_percent_change_from_baseline^2) +
             parks_percent_change_from_baseline + 
             transit_stations_percent_change_from_baseline +
             I(transit_stations_percent_change_from_baseline^2) +
             workplaces_percent_change_from_baseline +
             residential_percent_change_from_baseline +
             value +
             density +
             num_large_cities +
             `Black orAfrican American` +
             `American Indianand Alaska Native` +
             bSpline(time,df=5),
             offset = log(population), 
           data=train_set)
mod6_preds <- predict.lm(mod6,test_set)
plot(exp(mod6_preds),test_set$cases)
abline(0,1)
sqrt(mean(((exp(mod6_preds) - test_set$cases)^2), na.rm=TRUE))
# bad! abort!
```

```{r}
# remove workplace mobility 
mod7 <- lm(log(cases) ~ retail_and_recreation_percent_change_from_baseline + 
             grocery_and_pharmacy_percent_change_from_baseline + 
             I(grocery_and_pharmacy_percent_change_from_baseline^2) +
             parks_percent_change_from_baseline + 
             transit_stations_percent_change_from_baseline +
             residential_percent_change_from_baseline +
             value +
             density +
             num_large_cities +
             `Black orAfrican American` +
             `American Indianand Alaska Native` +
             bSpline(time,df=5),
             offset = log(population), 
           data=train_set)
mod7_preds <- predict.lm(mod7,test_set)
plot(exp(mod7_preds),test_set$cases)
abline(0,1)
sqrt(mean(((exp(mod7_preds) - test_set$cases)^2), na.rm=TRUE))
# good!
```

```{r}
# remove residential mobility
mod8 <- lm(log(cases) ~ retail_and_recreation_percent_change_from_baseline + 
             grocery_and_pharmacy_percent_change_from_baseline + 
             I(grocery_and_pharmacy_percent_change_from_baseline^2) +
             parks_percent_change_from_baseline + 
             transit_stations_percent_change_from_baseline +
             value +
             density +
             num_large_cities +
             `Black orAfrican American` +
             `American Indianand Alaska Native` +
             bSpline(time,df=5),
             offset = log(population), 
           data=train_set)
mod8_preds <- predict.lm(mod8,test_set)
plot(exp(mod8_preds),test_set$cases)
abline(0,1)
sqrt(mean(((exp(mod8_preds) - test_set$cases)^2), na.rm=TRUE))
# good!
```

```{r}
# add value (mask) quadratic term
mod9 <- lm(log(cases) ~ retail_and_recreation_percent_change_from_baseline + 
             grocery_and_pharmacy_percent_change_from_baseline + 
             I(grocery_and_pharmacy_percent_change_from_baseline^2) +
             parks_percent_change_from_baseline + 
             transit_stations_percent_change_from_baseline +
             value +
             I(value^2) +
             density +
             num_large_cities +
             `Black orAfrican American` +
             `American Indianand Alaska Native` +
             bSpline(time,df=5),
             offset = log(population), 
           data=train_set)
mod9_preds <- predict.lm(mod9,test_set)
plot(exp(mod9_preds),test_set$cases)
abline(0,1)
sqrt(mean(((exp(mod9_preds) - test_set$cases)^2), na.rm=TRUE))
# bad! abort!
```

```{r}
# remove transit mobility
mod10 <- lm(log(cases) ~ retail_and_recreation_percent_change_from_baseline + 
             grocery_and_pharmacy_percent_change_from_baseline + 
             I(grocery_and_pharmacy_percent_change_from_baseline^2) +
             parks_percent_change_from_baseline + 
             value +
             density +
             num_large_cities +
             `Black orAfrican American` +
             `American Indianand Alaska Native` +
             bSpline(time,df=5),
             offset = log(population), 
           data=train_set)
mod10_preds <- predict.lm(mod10,test_set)
plot(exp(mod10_preds),test_set$cases)
abline(0,1)
sqrt(mean(((exp(mod10_preds) - test_set$cases)^2), na.rm=TRUE))
```

```{r}
# take log of % Black or AA
mod11 <- lm(log(cases) ~ retail_and_recreation_percent_change_from_baseline + 
             grocery_and_pharmacy_percent_change_from_baseline + 
             I(grocery_and_pharmacy_percent_change_from_baseline^2) +
             parks_percent_change_from_baseline + 
             value +
             density +
             num_large_cities +
             log(`Black orAfrican American`) +
             `American Indianand Alaska Native` +
             bSpline(time,df=5),
             offset = log(population), 
           data=train_set)
mod11_preds <- predict.lm(mod11,test_set)
plot(exp(mod11_preds),test_set$cases)
abline(0,1)
sqrt(mean(((exp(mod11_preds) - test_set$cases)^2), na.rm=TRUE))
```

```{r}
# take log of % Am Indian/AK native
mod12 <- lm(log(cases) ~ retail_and_recreation_percent_change_from_baseline + 
             grocery_and_pharmacy_percent_change_from_baseline + 
             I(grocery_and_pharmacy_percent_change_from_baseline^2) +
             parks_percent_change_from_baseline + 
             value +
             density +
             num_large_cities +
             log(`Black orAfrican American`) +
             log(`American Indianand Alaska Native`) +
             bSpline(time,df=5),
             offset = log(population), 
           data=train_set)
mod12_preds <- predict.lm(mod12,test_set)
plot(exp(mod12_preds),test_set$cases)
abline(0,1)
sqrt(mean(((exp(mod12_preds) - test_set$cases)^2), na.rm=TRUE))
# bad! abort!
```

```{r}
# remove mask wearing value
mod13 <- lm(log(cases) ~ retail_and_recreation_percent_change_from_baseline + 
             grocery_and_pharmacy_percent_change_from_baseline + 
             I(grocery_and_pharmacy_percent_change_from_baseline^2) +
             parks_percent_change_from_baseline +
             num_large_cities +
             density + 
             log(`Black orAfrican American`) +
             `American Indianand Alaska Native` +
             bSpline(time,df=5),
             offset = log(population), 
           data=train_set)
mod13_preds <- predict.lm(mod13,test_set)
plot(exp(mod13_preds),test_set$cases)
abline(0,1)
sqrt(mean(((exp(mod13_preds) - test_set$cases)^2), na.rm=TRUE))
# better!
```

```{r}
# remove density 
mod14 <- lm(log(cases) ~ retail_and_recreation_percent_change_from_baseline + 
             grocery_and_pharmacy_percent_change_from_baseline + 
             I(grocery_and_pharmacy_percent_change_from_baseline^2) +
             parks_percent_change_from_baseline +
             num_large_cities +
             log(`Black orAfrican American`) +
             `American Indianand Alaska Native` +
             bSpline(time,df=5),
             offset = log(population), 
           data=train_set)
mod14_preds <- predict.lm(mod14,test_set)
plot(exp(mod14_preds),test_set$cases)
abline(0,1)
sqrt(mean(((exp(mod14_preds) - test_set$cases)^2), na.rm=TRUE))
# better!
```

```{r}
# remove % Am Indian/AK Native
mod15 <- lm(log(cases) ~ retail_and_recreation_percent_change_from_baseline + 
             grocery_and_pharmacy_percent_change_from_baseline + 
             I(grocery_and_pharmacy_percent_change_from_baseline^2) +
             parks_percent_change_from_baseline +
             num_large_cities +
             log(`Black orAfrican American`) +
             bSpline(time,df=5),
             offset = log(population), 
           data=train_set)
mod15_preds <- predict.lm(mod15,test_set)
plot(exp(mod15_preds),test_set$cases)
abline(0,1)
sqrt(mean(((exp(mod15_preds) - test_set$cases)^2), na.rm=TRUE))
# better!
```

```{r}
# remove quadratic term for grocery mobility
mod16 <- lm(log(cases) ~ retail_and_recreation_percent_change_from_baseline + 
             grocery_and_pharmacy_percent_change_from_baseline + 
             parks_percent_change_from_baseline +
             num_large_cities +
             log(`Black orAfrican American`) +
             bSpline(time,df=5),
             offset = log(population), 
           data=train_set)
mod16_preds <- predict.lm(mod16,test_set)
plot(exp(mod16_preds),test_set$cases)
abline(0,1)
sqrt(mean(((exp(mod16_preds) - test_set$cases)^2), na.rm=TRUE))
```

```{r}
# remove retail mobility 
mod17 <- lm(log(cases) ~  
             grocery_and_pharmacy_percent_change_from_baseline + 
             parks_percent_change_from_baseline +
             num_large_cities +
             log(`Black orAfrican American`) +
             bSpline(time,df=5),
             offset = log(population), 
           data=train_set)
mod17_preds <- predict.lm(mod17,test_set)
plot(exp(mod17_preds),test_set$cases)
abline(0,1)
sqrt(mean(((exp(mod17_preds) - test_set$cases)^2), na.rm=TRUE))
```

```{r}
# remove grocery mobility
mod18 <- lm(log(cases) ~  
             parks_percent_change_from_baseline +
             num_large_cities +
             log(`Black orAfrican American`) +
             bSpline(time,df=5),
             offset = log(population), 
           data=train_set)
mod18_preds <- predict.lm(mod18,test_set)
plot(exp(mod18_preds),test_set$cases)
abline(0,1)
sqrt(mean(((exp(mod18_preds) - test_set$cases)^2), na.rm=TRUE))
```


Why is mask wearing not useful for prediction?
```{r}
library(maps)
states <- map_data("state")
```

```{r}
final_df$state <- tolower(final_df$state)
colnames(states)[5] <- "state"
map_df <- left_join(final_df,states,by="state")
```

```{r}
map_df %>% filter(time==73) %>% ggplot(aes(x = long, y = lat, group = group)) +
   geom_polygon(aes(fill=value), color = "white") + 
   theme(panel.grid.major = element_blank(), 
         panel.background = element_blank(),
         axis.title = element_blank(), 
         axis.text = element_blank(),
         axis.ticks = element_blank()) +
   ggtitle("Percent of People Reporting They Wear a Mask in Public on 11/19/20") + 
   scale_fill_viridis_c(name="")
```

```{r}
final_df %>% filter(time==73) %>% ggplot(aes(x=state,y=cases)) + geom_bar(stat="identity") + theme(axis.text.x = element_text(angle = 90))
```

```{r}
ggplot(final_df,aes(x=value,y=new_cases)) + geom_point()
```

When it comes to prediction, variables related to population size and population density are more useful than the CMU mask data.




EXTRA CODE, UNNEEDED
trying a first attempt at regression now, mask data not looking super usable

need to update date variables to be time (day = 1 to day = whatever)


```{r}
library(splines)
library(splines2)
mod1 <- lm(log(cases) ~ retail_and_recreation_percent_change_from_baseline + 
             grocery_and_pharmacy_percent_change_from_baseline + 
             parks_percent_change_from_baseline + 
             transit_stations_percent_change_from_baseline +
             workplaces_percent_change_from_baseline +
             gov_party + 
             num_large_cities +
             `Black orAfrican American` +
             bSpline(time,knots=5),
             offset = log(population), 
             data=final_df
             )
summary(mod1)
```



Interesting that cases decreased in states where there is, going to add population density before splitting into training/testing sets because that could be a confounder.



```{r}
plot(final_df2$date,final_df2$cases)
```

```{r}
scatter.smooth(final_df2$grocery_and_pharmacy_percent_change_from_baseline,log(final_df2$cases),col="pink")
```

```{r}
scatter.smooth(final_df2$parks_percent_change_from_baseline,log(final_df2$cases),col="pink")
```

```{r}
scatter.smooth(final_df2$transit_stations_percent_change_from_baseline,log(final_df2$cases),col="pink")
```

```{r}
scatter.smooth(final_df2$workplaces_percent_change_from_baseline,log(final_df2$cases),col="pink")
```

```{r}
plot(as.factor(final_df2$num_large_cities),log(final_df2$cases))
```

```{r}
plot(final_df2$gov_party,log(final_df2$cases))
```

```{r}
scatter.smooth(log(final_df2$`Black orAfrican American`),log(final_df2$cases),col="pink")
```

```{r}
scatter.smooth(log(final_df2$density),log(final_df2$cases),col="pink")
```


Split into Training/Testing Sets:
```{r}
final_df2 <- final_df2 %>% arrange(date)
train_set <- final_df2[1:112351,]
test_set <- final_df2[112351:127050,]
```

```{r}
mod2 <- lm(log(cases) ~ retail_and_recreation_percent_change_from_baseline + 
             grocery_and_pharmacy_percent_change_from_baseline + 
             parks_percent_change_from_baseline + 
             transit_stations_percent_change_from_baseline +
             workplaces_percent_change_from_baseline +
             gov_party + 
             num_large_cities +
             `Black orAfrican American` +
             density +
             bSpline(time,df=10),
             offset = log(population), 
             data=train_set
             )
summary(mod2)
```

```{r}
mod2_preds <- predict.lm(mod2,test_set)
plot(exp(mod2_preds),test_set$cases)
abline(0,1)
sqrt(mean((test_set$cases - exp(mod2_preds))^2))
```

Note: it makes sense that this is so bad for prediction because our test set is the month of November, and there was a massive spike in November.

```{r}
mod3 <- lm(log(cases) ~ retail_and_recreation_percent_change_from_baseline + 
             grocery_and_pharmacy_percent_change_from_baseline + 
             parks_percent_change_from_baseline + 
             transit_stations_percent_change_from_baseline +
             workplaces_percent_change_from_baseline +
             gov_party + 
             num_large_cities +
             `Black orAfrican American` +
             density +
             bSpline(time,df=12),
             offset = log(population), 
             data=train_set
             )
mod3_preds <- predict.lm(mod3,test_set)
plot(exp(mod3_preds),test_set$cases)
abline(0,1)
sqrt(mean((test_set$cases - exp(mod3_preds))^2))
```

```{r}
mod4 <- lm(log(cases) ~ retail_and_recreation_percent_change_from_baseline + 
             grocery_and_pharmacy_percent_change_from_baseline + 
             parks_percent_change_from_baseline + 
             transit_stations_percent_change_from_baseline +
             workplaces_percent_change_from_baseline +
             gov_party + 
             num_large_cities +
             `Black orAfrican American` +
             density +
             bSpline(time,df=15),
             offset = log(population), 
             data=train_set
             )
mod4_preds <- predict.lm(mod4,test_set)
plot(exp(mod4_preds),test_set$cases)
abline(0,1)
sqrt(mean((test_set$cases - exp(mod4_preds))^2))
```

```{r}
mod5 <- lm(log(cases) ~ retail_and_recreation_percent_change_from_baseline + 
             grocery_and_pharmacy_percent_change_from_baseline + 
             parks_percent_change_from_baseline + 
             transit_stations_percent_change_from_baseline +
             workplaces_percent_change_from_baseline +
             gov_party + 
             num_large_cities +
             `Black orAfrican American` +
             density +
             bSpline(time,df=18),
             offset = log(population), 
             data=train_set
             )
mod5_preds <- predict.lm(mod5,test_set)
plot(exp(mod5_preds),test_set$cases)
abline(0,1)
sqrt(mean((test_set$cases - exp(mod5_preds))^2))
```

```{r}
mod6 <- lm(log(cases) ~ retail_and_recreation_percent_change_from_baseline + 
             grocery_and_pharmacy_percent_change_from_baseline + 
             parks_percent_change_from_baseline + 
             transit_stations_percent_change_from_baseline +
             workplaces_percent_change_from_baseline +
             gov_party + 
             num_large_cities +
             `Black orAfrican American` +
             density +
             bSpline(time,df=17),
             offset = log(population), 
             data=train_set
             )
mod6_preds <- predict.lm(mod6,test_set)
plot(exp(mod6_preds),test_set$cases)
abline(0,1)
sqrt(mean((test_set$cases - exp(mod6_preds))^2))
```

```{r}
mod7 <- lm(log(cases) ~ retail_and_recreation_percent_change_from_baseline + 
             grocery_and_pharmacy_percent_change_from_baseline + 
             parks_percent_change_from_baseline + 
             transit_stations_percent_change_from_baseline +
             workplaces_percent_change_from_baseline +
             gov_party + 
             num_large_cities +
             `Black orAfrican American` +
             density +
             bSpline(time,df=16),
             offset = log(population), 
             data=train_set
             )
mod7_preds <- predict.lm(mod7,test_set)
plot(exp(mod7_preds),test_set$cases)
abline(0,1)
sqrt(mean((test_set$cases - exp(mod7_preds))^2))
```

```{r}
mod8 <- lm(log(cases) ~ retail_and_recreation_percent_change_from_baseline + 
             grocery_and_pharmacy_percent_change_from_baseline + 
             parks_percent_change_from_baseline + 
             transit_stations_percent_change_from_baseline +
             workplaces_percent_change_from_baseline +
             gov_party + 
             num_large_cities +
             `Black orAfrican American` +
             density +
             bSpline(time,df=14),
             offset = log(population), 
             data=train_set
             )
mod8_preds <- predict.lm(mod8,test_set)
plot(exp(mod8_preds),test_set$cases)
abline(0,1)
sqrt(mean((test_set$cases - exp(mod8_preds))^2))
```

```{r}
mod9 <- lm(log(cases) ~ retail_and_recreation_percent_change_from_baseline + 
             grocery_and_pharmacy_percent_change_from_baseline + 
             parks_percent_change_from_baseline + 
             transit_stations_percent_change_from_baseline +
             workplaces_percent_change_from_baseline +
             gov_party + 
             num_large_cities +
             `Black orAfrican American` +
             log(density) +
             bSpline(time,df=15),
             offset = log(population), 
             data=train_set
             )
mod9_preds <- predict.lm(mod9,test_set)
plot(exp(mod9_preds),test_set$cases)
abline(0,1)
sqrt(mean((test_set$cases - exp(mod9_preds))^2))
```

```{r}
mod10 <- lm(log(cases) ~ retail_and_recreation_percent_change_from_baseline + 
             grocery_and_pharmacy_percent_change_from_baseline + 
             parks_percent_change_from_baseline + 
             transit_stations_percent_change_from_baseline +
             workplaces_percent_change_from_baseline +
             gov_party + 
             num_large_cities +
             log(`Black orAfrican American`) +
             density +
             bSpline(time,df=15),
             offset = log(population), 
             data=train_set
             )
mod10_preds <- predict.lm(mod10,test_set)
plot(exp(mod10_preds),test_set$cases)
abline(0,1)
sqrt(mean((test_set$cases - exp(mod10_preds))^2))
```

```{r}
mod11 <- lm(log(cases) ~  retail_and_recreation_percent_change_from_baseline +
             I(retail_and_recreation_percent_change_from_baseline^2) +
             grocery_and_pharmacy_percent_change_from_baseline + 
             parks_percent_change_from_baseline + 
             transit_stations_percent_change_from_baseline +
             workplaces_percent_change_from_baseline +
             gov_party + 
             num_large_cities +
             log(`Black orAfrican American`) +
             density +
             bSpline(time,df=15),
             offset = log(population), 
             data=train_set
             )
mod11_preds <- predict.lm(mod11,test_set)
plot(exp(mod11_preds),test_set$cases)
abline(0,1)
sqrt(mean((test_set$cases - exp(mod11_preds))^2))
```

```{r}
mod12 <- lm(log(cases) ~  
             grocery_and_pharmacy_percent_change_from_baseline + 
             parks_percent_change_from_baseline + 
             transit_stations_percent_change_from_baseline +
             workplaces_percent_change_from_baseline +
             gov_party + 
             num_large_cities +
             log(`Black orAfrican American`) +
             density +
             bSpline(time,df=15),
             offset = log(population), 
             data=train_set
             )
mod12_preds <- predict.lm(mod12,test_set)
plot(exp(mod12_preds),test_set$cases)
abline(0,1)
sqrt(mean((test_set$cases - exp(mod12_preds))^2))
```
```{r}
mod13 <- lm(log(cases) ~  
             parks_percent_change_from_baseline + 
             transit_stations_percent_change_from_baseline +
             workplaces_percent_change_from_baseline +
             gov_party + 
             num_large_cities +
             log(`Black orAfrican American`) +
             density +
             bSpline(time,df=15),
             offset = log(population), 
             data=train_set
             )
mod13_preds <- predict.lm(mod13,test_set)
plot(exp(mod13_preds),test_set$cases)
abline(0,1)
sqrt(mean((test_set$cases - exp(mod13_preds))^2))
```

```{r}
mod14 <- lm(log(cases) ~  
             transit_stations_percent_change_from_baseline +
             workplaces_percent_change_from_baseline +
             gov_party + 
             num_large_cities +
             log(`Black orAfrican American`) +
             density +
             bSpline(time,df=15),
             offset = log(population), 
             data=train_set
             )
mod14_preds <- predict.lm(mod14,test_set)
plot(exp(mod14_preds),test_set$cases)
abline(0,1)
sqrt(mean((test_set$cases - exp(mod14_preds))^2))
```

```{r}
mod15 <- lm(log(cases) ~  
             parks_percent_change_from_baseline + 
             workplaces_percent_change_from_baseline +
             gov_party + 
             num_large_cities +
             log(`Black orAfrican American`) +
             density +
             bSpline(time,df=15),
             offset = log(population), 
             data=train_set
             )
mod15_preds <- predict.lm(mod15,test_set)
plot(exp(mod15_preds),test_set$cases)
abline(0,1)
sqrt(mean((test_set$cases - exp(mod15_preds))^2))
```

```{r}
mod16 <- lm(log(cases) ~  
             parks_percent_change_from_baseline + 
             transit_stations_percent_change_from_baseline +
             gov_party + 
             num_large_cities +
             log(`Black orAfrican American`) +
             density +
             bSpline(time,df=15),
             offset = log(population), 
             data=train_set
             )
mod16_preds <- predict.lm(mod16,test_set)
plot(exp(mod16_preds),test_set$cases)
abline(0,1)
sqrt(mean((test_set$cases - exp(mod16_preds))^2))
```

Maybe try to model new cases instead of total, or maybe the issue is that the states just behave too differently and creating one model for the entire country is unrealistic

testing making training set 80% of obs per state, NOT based on time:
```{r}
train_index <- createDataPartition(final_df2$state, times = 1, p = 0.9, list = FALSE)
```

```{r}
train_set2 <- final_df2[train_index, ]
test_set2 <- final_df2[-train_index, ]
```

```{r}
mod17 <- lm(log(cases) ~  
              grocery_and_pharmacy_percent_change_from_baseline + 
              log(`Black orAfrican American`) +
             bSpline(time,df=10),
             offset = log(population), 
             data=train_set2
             )
mod17_preds <- predict.lm(mod17,test_set2)
plot(exp(mod17_preds),test_set2$cases)
abline(0,1)
sqrt(mean((test_set2$cases - exp(mod17_preds))^2))
```

